---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

In this report we analyze data from a personal activity monitoring device as described [here](README.md).

## Loading and preprocessing the data

We will use the `dplyr` pckage to manipulate the data, so load it first.

```{r}
library(dplyr)
```

We load the data directly from the zip file, and turn it into a `tbl`.

```{r}
activity <- tbl_df(read.csv(file = unz(description = "activity.zip", filename = "activity.csv")))
activity
```

## What is mean total number of steps taken per day?

We calculate the total number of steps taken per day, ignoring missing values.

```{r}
by_day <- activity %>% filter(!is.na(steps)) %>% group_by(date) %>% summarize(steps = sum(steps))
by_day
```

We make a histogram of the total number of steps taken each day.

```{r}
library(ggplot2)
ggplot(data = by_day, mapping = aes(x = steps)) + geom_histogram()
```

We calculate the mean and median of the total number of steps taken each day.

```{r}
summ <- by_day %>% summarize(mean = mean(steps), median = median(steps))
summ
```

So the mean number of steps taken per day is `r format(summ$mean)` and the median is `r summ$median`.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
