---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

In this report we analyze data from a personal activity monitoring device as described [here](README.md).

## Loading and preprocessing the data

We will use the `dplyr` pckage to manipulate the data, so load it first.

```{r}
library(dplyr)
```

We load the data directly from the zip file, and turn it into a `tbl`.

```{r}
activity <- tbl_df(read.csv(file = unz(description = "activity.zip", filename = "activity.csv")))
activity
```

## What is mean total number of steps taken per day?

We calculate the total number of steps taken per day, ignoring missing values.

```{r}
by_date <- activity %>% filter(!is.na(steps)) %>% group_by(date) %>% summarize(steps = sum(steps))
by_date
```

We make a histogram of the total number of steps taken each day.

```{r}
library(ggplot2)
ggplot(data = by_date, mapping = aes(x = steps)) +
  geom_histogram() +
  xlab("Steps taken") +
  ylab("Count") +
  ggtitle("Total number of steps taken per day")
```

We calculate the mean and median of the total number of steps taken each day.

```{r}
summ <- by_date %>% summarize(mean = mean(steps), median = median(steps))
summ
```

So the mean number of steps taken per day is `r format(summ$mean)` and the median is `r summ$median`.

## What is the average daily activity pattern?

We make a time series plot of the average number of steps taken for each time interval. First we need to find the mean number of steps for each interval, again ignoring missing data. Note that the `mutate` call derives from the interval format a simple number of minutes after midnight.

```{r}
by_interval <-
  activity %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps)) %>%
  mutate(interval_mins = 60 * (interval %/% 100) + interval %% 100)
```

Then we make the plot.

```{r}
ggplot(data = by_interval, mapping = aes(x = interval_mins, y = steps)) +
  geom_line() +
  xlab("Interval") +
  ylab("Number of steps") +
  ggtitle("Average daily activity pattern")
```

Which of these intervals contains the maximum number of steps?

```{r}
by_interval$interval[which.max(by_interval$steps)]
```

This corresponds with the interval at 8:35 am.

## Imputing missing values

We determine how many missing values are in the dataset.

```{r}
sum(is.na(activity$steps))
```

Since we expect the daily activity pattern to be reasonable regular, it makes sense to impute missing values based on the mean for the same interval on other days. We fill in the missing data based on this strategy.

```{r}
imputed <- tbl_df(merge(x = activity, y = by_interval, by = "interval")) %>%
  mutate(steps = ifelse(is.na(steps.x), steps.y, steps.x))
imputed
```

Note that in the resulting dataset, `steps.x` is the original observed value (which might be `NA`); `steps.y` is the mean value that will be used for imputation if necessary; and `steps` is the final (possibly imputed) value.

Next we generate another histogram using the imputed values.

```{r}
by_date_imputed <- imputed %>% group_by(date) %>% summarize(steps = sum(steps))
ggplot(data = by_date_imputed, mapping = aes(x = steps)) +
  geom_histogram() +
  xlab("Steps taken") +
  ylab("Count") +
  ggtitle("Total number of steps taken per day")
```

We recalculate the mean and median of the total number of steps taken each day.

```{r}
summ_imputed <- by_date_imputed %>% summarize(mean = mean(steps), median = median(steps))
summ_imputed
```

So the mean number of steps taken per day is `r format(summ_imputed$mean)` and the median is `r format(summ_imputed$median)`. As we would expect, imputation does not significantly affect the mean or median.

## Are there differences in activity patterns between weekdays and weekends?

We add a `type` variable that tells whether the date is a weekday or a weekend day.

Note that we use the `lubridate::wday` instead of `base::weekdays`. `base::weekdays`` returns a locale-specific string, so comparisons based on this value will fail if the locale changes. On the other hand, `lubridate::wday` by default returns a numeric value that is independant of locale.

```{r}
library(lubridate)
with_type <- imputed %>% mutate(type = factor(wday(date) %in% c(1, 7),
                                              levels = c(TRUE, FALSE),
                                              labels = c("weekend", "weekday")))
with_type
```

Next we make a time series plot.

```{r}
by_interval_and_type <-
  with_type %>%
  group_by(type, interval_mins) %>%
  summarize(steps = mean(steps))
ggplot(data = by_interval_and_type, mapping = aes(x = interval_mins, y = steps)) +
  geom_line() +
  facet_grid(facets = type ~ .) +
  xlab("Interval") +
  ylab("Number of steps") +
  ggtitle("Average daily activity pattern (weekend vs weekday)")
```
